---
title: "D498A_final proj_v1"
author: "Damon Shorty"
date: "4/4/2023"
output: pdf_document
---

# libraries 
```{r}
# library(corrplot)
library(ggplot2)
# library(MASS) #for LDA, QDA
library(class) #for KNN
library(e1071) #for Naive Bayes, SVM
library(leaps) #for Regularization, regsubsets
library(tree)
library(randomForest)
```




# read data  
```{r}
# primary <- read.csv(file = 'primary_data.csv')
mushrooms <- read.csv(file = 'secondary_data2.csv',stringsAsFactors = TRUE)

```

```{r}
# str(mushrooms)
```


# check data  
```{r}
#check for NAs
sum(is.na(mushrooms))
```


# clean up data 
```{r}
#edit 1
levels(mushrooms$cap.surface)[1] <- "v"

#edit 2
levels(mushrooms$gill.attachment)[1] <- "v"
 
#edit 3
levels(mushrooms$gill.spacing)[1] <- "v"
 
#edit 4
levels(mushrooms$stem.root)[1] <- "v"
 
#edit 5
levels(mushrooms$stem.surface)[1] <- "v"
 
#edit 6
levels(mushrooms$veil.type)[1] <- "p"
 
#edit 7
levels(mushrooms$veil.color)[1] <- "v"

#edit 8
levels(mushrooms$ring.type)[1] <- "v"

#edit 9
levels(mushrooms$spore.print.color)[1] <- "v"

```




```{r}
# str(mushrooms) #for check only
```



# Summary Statistics   

```{r}
summary(mushrooms)
```

```{r}
# head(mushrooms)
```





## quantiative vars. 

```{r}
pairs(mushrooms[,c(1,2,10,11)])
```




```{r}
round(cor(mushrooms[,c(2,10,11)]),3)
```






## qualitative vars. 

```{r}
ggplot(mushrooms,aes(x=cap.surface,y=cap.shape,col=class)) +
  geom_jitter() +
  ggtitle("Cap Surface vs. Cap Shape") +
  xlab("Cap Surface Type") +
  ylab("Cap Shape Type") +
  theme(plot.title = element_text(face="bold")) +
  scale_color_discrete(labels=c('Edible','Poisonous')) +
  scale_y_discrete(labels=c("bell","conical","flat","others","spherical","sunken","convex")) +
  scale_x_discrete(labels=c("unknown","unknown","fleshy","grooves","shiny","fibrous","silky","leathery","smooth","sticky","wrinkled","scaly")) +
  theme(axis.text.x = element_text(angle = 90,hjust=1,vjust=0))


ggplot(mushrooms,aes(x=cap.color,y=cap.shape,col=class)) +
  geom_jitter() +
  ggtitle("Cap Color vs. Cap Shape") +
  xlab("Cap Color") +
  ylab("Cap Shape Type") +
  theme(plot.title = element_text(face="bold")) +
  scale_color_discrete(labels=c('Edible','Poisonous')) +
  scale_y_discrete(labels=c("bell","conical","flat","others","spherical","sunken","convex")) +
  scale_x_discrete(labels=c("buff","red","gray","black","blue","brown","orange","pink","green","purple","white","yellow")) +
  theme(axis.text.x = element_text(angle = 90,hjust=1,vjust=0))


ggplot(mushrooms,aes(x=does.bruise.or.bleed,y=cap.shape,col=class)) +
  geom_jitter() +
  ggtitle("Does Bruise/Bleed vs. Cap Shape") +
  xlab("Does Bruise or Bleed") +
  ylab("Cap Shape Type") +
  theme(plot.title = element_text(face="bold")) +
  scale_color_discrete(labels=c('Edible','Poisonous')) +
  scale_y_discrete(labels=c("bell","conical","flat","others","spherical","sunken","convex")) +
  scale_x_discrete(labels=c("no","yes"))


ggplot(mushrooms,aes(x=gill.attachment,y=cap.shape,col=class)) +
  geom_jitter() +
  ggtitle("Gill Attachment vs. Cap Shape") +
  xlab("Gill Attachment Type") +
  ylab("Cap Shape Type") +
  theme(plot.title = element_text(face="bold")) +
  scale_color_discrete(labels=c('Edible','Poisonous')) +
  scale_y_discrete(labels=c("bell","conical","flat","others","spherical","sunken","convex")) +
  scale_x_discrete(labels=c("unknown","adnate","decurrent","free","none","pores","sinute","adnexed")) +
  theme(axis.text.x = element_text(angle = 90,hjust=1,vjust=0))


ggplot(mushrooms,aes(x=gill.spacing,y=cap.shape,col=class)) +
  geom_jitter() +
  ggtitle("Gill Spacing vs. Cap Shape") +
  xlab("Gill Spacing") +
  ylab("Cap Shape Type") +
  theme(plot.title = element_text(face="bold")) +
  scale_color_discrete(labels=c('Edible','Poisonous')) +
  scale_y_discrete(labels=c("bell","conical","flat","others","spherical","sunken","convex")) +
  scale_x_discrete(labels=c("unknown","close","distant","none"))


ggplot(mushrooms,aes(x=gill.color,y=cap.shape,col=class)) +
  geom_jitter() +
  ggtitle("Gill Color vs. Cap Shape") +
  xlab("Gill Color") +
  ylab("Cap Shape Type") +
  theme(plot.title = element_text(face="bold")) +
  scale_color_discrete(labels=c('Edible','Poisonous')) +
  scale_y_discrete(labels=c("bell","conical","flat","others","spherical","sunken","convex")) +
  scale_x_discrete(labels=c("buff","red","none","gray","black","brown","orange","pink","green","purple","white","yellow")) +
  theme(axis.text.x = element_text(angle = 90,hjust=1,vjust=0))


ggplot(mushrooms,aes(x=stem.root,y=cap.shape,col=class)) +
  geom_jitter() +
  ggtitle("Stem Root vs. Cap Shape") +
  xlab("Stem Root Type") +
  ylab("Cap Shape Type") +
  theme(plot.title = element_text(face="bold")) +
  scale_color_discrete(labels=c('Edible','Poisonous')) +
  scale_y_discrete(labels=c("bell","conical","flat","others","spherical","sunken","convex")) +
  scale_x_discrete(labels=c("unknown","bulbous","club","unknown","rooted","swollen"))


ggplot(mushrooms,aes(x=stem.surface,y=cap.shape,col=class)) +
  geom_jitter() +
  ggtitle("Stem Surface vs. Cap Shape") +
  xlab("Stem Surface Type") +
  ylab("Cap Shape Type") +
  theme(plot.title = element_text(face="bold")) +
  scale_color_discrete(labels=c('Edible','Poisonous')) +
  scale_y_discrete(labels=c("bell","conical","flat","others","spherical","sunken","convex")) +
  scale_x_discrete(labels=c("unknown","none","grooves","shiny","fibrous","silky","smooth","sticky","scaly")) +
  theme(axis.text.x = element_text(angle = 90,hjust=1,vjust=0))


ggplot(mushrooms,aes(x=stem.color,y=cap.shape,col=class)) +
  geom_jitter() +
  ggtitle("Stem Color vs. Cap Shape") +
  xlab("Stem Color") +
  ylab("Cap Shape Type") +
  theme(plot.title = element_text(face="bold")) +
  scale_color_discrete(labels=c('Edible','Poisonous')) +
  scale_y_discrete(labels=c("bell","conical","flat","others","spherical","sunken","convex")) +
  scale_x_discrete(labels=c("buff","red","none","gray","black","blue","brown","orange","pink","green","purple","white","yellow")) +
  theme(axis.text.x = element_text(angle = 90,hjust=1,vjust=0))


ggplot(mushrooms,aes(x=veil.type,y=cap.shape,col=class)) +
  geom_jitter() +
  ggtitle("Veil Type vs. Cap Shape") +
  xlab("Veil Type") +
  ylab("Cap Shape Type") +
  theme(plot.title = element_text(face="bold")) +
  scale_color_discrete(labels=c('Edible','Poisonous')) +
  scale_y_discrete(labels=c("bell","conical","flat","others","spherical","sunken","convex")) +
  scale_x_discrete(labels=c("partial","universal"))


ggplot(mushrooms,aes(x=veil.color,y=cap.shape,col=class)) +
  geom_jitter() +
  ggtitle("Veil Color vs. Cap Shape") +
  xlab("Veil Color") +
  ylab("Cap Shape Type") +
  theme(plot.title = element_text(face="bold")) +
  scale_color_discrete(labels=c('Edible','Poisonous')) +
  scale_y_discrete(labels=c("bell","conical","flat","others","spherical","sunken","convex")) +
  scale_x_discrete(labels=c("unknown","red","black","brown","purple","white","yellow"))
  # theme(axis.text.x = element_text(angle = 90,hjust=1,vjust=0))


ggplot(mushrooms,aes(x=has.ring,y=cap.shape,col=class)) +
  geom_jitter() +
  ggtitle("Has Ring vs. Cap Shape") +
  xlab("Has Ring") +
  ylab("Cap Shape Type") +
  theme(plot.title = element_text(face="bold")) +
  scale_color_discrete(labels=c('Edible','Poisonous')) +
  scale_y_discrete(labels=c("bell","conical","flat","others","spherical","sunken","convex")) +
  scale_x_discrete(labels=c("no","yes"))


ggplot(mushrooms,aes(x=ring.type,y=cap.shape,col=class)) +
  geom_jitter() +
  ggtitle("Ring Type vs. Cap Shape") +
  xlab("Ring Type") +
  ylab("Cap Shape Type") +
  theme(plot.title = element_text(face="bold")) +
  scale_color_discrete(labels=c('Edible','Poisonous')) +
  scale_y_discrete(labels=c("bell","conical","flat","others","spherical","sunken","convex")) +
  scale_x_discrete(labels=c("unknown","evanescent","none","grooved","large","moveable","pendant","flaring","zone")) +
  theme(axis.text.x = element_text(angle = 90,hjust=1,vjust=0))


ggplot(mushrooms,aes(x=spore.print.color,y=cap.shape,col=class)) +
  geom_jitter() +
  ggtitle("Spore Print Color vs. Cap Shape") +
  xlab("Spore Print Color") +
  ylab("Cap Shape Type") +
  theme(plot.title = element_text(face="bold")) +
  scale_color_discrete(labels=c('Edible','Poisonous')) +
  scale_y_discrete(labels=c("bell","conical","flat","others","spherical","sunken","convex")) +
  scale_x_discrete(labels=c("unknown","gray","black","brown","pink","green","purple","white"))
  # theme(axis.text.x = element_text(angle = 90,hjust=1,vjust=0))



ggplot(mushrooms,aes(x=habitat,y=cap.shape,col=class)) +
  geom_jitter() +
  ggtitle("Habitat vs. Cap Shape") +
  xlab("Habitat") +
  ylab("Cap Shape Type") +
  theme(plot.title = element_text(face="bold")) +
  scale_color_discrete(labels=c('Edible','Poisonous')) +
  scale_y_discrete(labels=c("bell","conical","flat","others","spherical","sunken","convex")) +
  scale_x_discrete(labels=c("woods","grasses","heaths","leaves","meadows","paths","urban","waste"))



ggplot(mushrooms,aes(x=season,y=cap.shape,col=class)) +
  geom_jitter() +
  ggtitle("Season vs. Cap Shape") +
  xlab("Seasons") +
  ylab("Cap Shape Type") +
  theme(plot.title = element_text(face="bold")) +
  scale_color_discrete(labels=c('Edible','Poisonous')) +
  scale_y_discrete(labels=c("bell","conical","flat","others","spherical","sunken","convex")) +
  scale_x_discrete(labels=c("autumn","spring","summer","winter"))





```




















# ANALYSIS SECTION: 

## train & test data 
```{r}
set.seed(12)

train <- sample(1:nrow(mushrooms),0.75*nrow(mushrooms))


mushrooms_train <- mushrooms[train,]
mushrooms_test <- mushrooms[-train,]

nrow(mushrooms_train)+nrow(mushrooms_test)

```












# TREE based method 

### fitting classification trees 
```{r}
tree_mushrooms <- tree(class~.,data = mushrooms_train)

summary(tree_mushrooms)

```

number of terminal nodes = 28
training error rate = 0.08956


```{r}
plot(tree_mushrooms)
text(tree_mushrooms,pretty=0)
title(main="Unpruned Classification Tree")

```



```{r}
tree_mushrooms
```





```{r}
tree_pred <- predict(tree_mushrooms,mushrooms_test,type="class")

table(tree_pred,mushrooms_test$class)
```

```{r}
#correct predictions in the test data set
(6308+7622)/nrow(mushrooms_test)

#wrong pred.
1-((6308+7622)/nrow(mushrooms_test))

```

91.24% of the test observations are correctly classified (test error rate = 0.08763) 




### prune tree 


```{r}
cv_mushrooms <- cv.tree(tree_mushrooms, FUN=prune.misclass)
# names(cv_mushrooms)
cv_mushrooms

```



```{r}
# par(mfrow=c(1,2))
plot(cv_mushrooms$size,cv_mushrooms$dev,type="b",xlab = "Number of Trees",ylab="Test Classification Error")
# plot(cv_mushrooms$k,cv_mushrooms$dev,type="b")

```


```{r}
prune_mushrooms_misclass <- prune.misclass(tree_mushrooms,best=28)
plot(prune_mushrooms_misclass)
text(prune_mushrooms_misclass,pretty=0)
title(main="Pruned Classification Tree")

```


```{r}
tree_pred2 <- predict(prune_mushrooms_misclass,mushrooms_test,type="class")
table(tree_pred2,mushrooms_test$class)
```

```{r}
#the test observations that are correctly classified
(6308+7622)/nrow(mushrooms_test)

#test error rate
1-((6308+7622)/nrow(mushrooms_test))
```
91.24% of the test observations are correctly classified (test error rate = 0.08763)


















## Random Forest 

### bagging 

```{r}
set.seed(54)

bag_mushrooms <- randomForest(class~.,data = mushrooms_train,mtry=20,importance=TRUE) #mtry=20 indicates all 20 predictors should be considered
bag_mushrooms

```


```{r}
(20353+25439)/nrow(mushrooms_train)

#training error rate
1-((20353+25439)/nrow(mushrooms_train))
```

99.98% of the training observations are correctly classified (training error rate = 0.0002)



```{r}
plot(bag_mushrooms)
```



```{r}
tree_pred3 <- predict(bag_mushrooms,mushrooms_test,type="class")
table(tree_pred3,mushrooms_test$class)
```

```{r}
(6825+8440)/nrow(mushrooms_test)

#test error rate
1-((6825+8440)/nrow(mushrooms_test))
```

99.98% of the test observations are correctly classified (test error rate = 0.0002)












```{r}
set.seed(73)

bag_mushrooms <- randomForest(class~.,data = mushrooms_train,mtry=20,ntree=28) #28 number of trees
bag_mushrooms


```

```{r}
plot(bag_mushrooms)
```

```{r}
tree_pred4 <- predict(bag_mushrooms,mushrooms_train,type="class")
table(tree_pred4,mushrooms_train$class)
```

```{r}
(20356+25445)/nrow(mushrooms_train)
```





### rf 


```{r}
set.seed(28)

m <- round(sqrt(20),0)

rf_mushrooms <- randomForest(class~.,data = mushrooms_train,mtry=m,importance=TRUE)
rf_mushrooms
```

training error rate = 0



```{r}
tree_pred5 <- predict(rf_mushrooms,mushrooms_test,type="class")
table(tree_pred5,mushrooms_test$class)
```

```{r}
(6825+8443)/nrow(mushrooms_test)
```




```{r}
importance(rf_mushrooms)
```


```{r}
varImpPlot(rf_mushrooms,main="Variable Importance")
```






## Naive Bayes 

```{r}
nb_fit <- naiveBayes(class~.,data = mushrooms_train)
# nb_fit
```


```{r}
nb_class <- predict(nb_fit,mushrooms_test)
table(nb_class,mushrooms_test$class)
```

```{r}
mean(nb_class==mushrooms_test$class)

#test error rate
1-(mean(nb_class==mushrooms_test$class))
```

74.28% of test observations are correctly classified (test error rate = 0.25720) 










# Support Vector Machine Classifier 


```{r}
svm_fit <- svm(class~.,data = mushrooms_train,kernel="radial",gamma=1,cost=1)
```

```{r}
summary(svm_fit)
```

number of support vectors = 10274
(5557,4747)


```{r}
tree_pred6 <- predict(svm_fit,mushrooms_test,type="class")
table(tree_pred6,mushrooms_test$class)
```


```{r}
(6823+8443)/nrow(mushrooms_test)

#test error rate
1-((6823+8443)/nrow(mushrooms_test))
```

99.99% test observations are correctly classified (test error rate = 0.00013)






## tune out


```{r}
#NOTICE/CAUTION: this will RUN for APPROX. 2 HOURS 15 MINS plus (depending on PC)

tune_out1 <- tune(svm,class~.,data = mushrooms_train,kernel="radial",cost=c(0.01,0.1,1,5,10))
```

```{r}
summary(tune_out1)
tune_out1$best.performance
```






```{r}
#NOTICE/CAUTION: this will RUN for APPROX. 4 HOURS 30 MINS plus (depending on PC)

tune_out2 <- tune(svm,class~.,data = mushrooms_train,kernel="radial",ranges = list(cost=c(0.01,0.1,1,5),gamma=1))
```




```{r}
summary(tune_out2)
```



```{r}
table(true=mushrooms_test$class,pred=predict(tune_out2$best.model,newdata = mushrooms_test))
```

```{r}
(6823+8443)/nrow(mushrooms_test)

#test error rate
1-((6823+8443)/nrow(mushrooms_test))
```

test error rate = 0.00013 




